---
title: "HW5 - Make Data Move"
author: "Agnes Boel Nielsen"
date: "05/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Explore global development with R
Today, you will load a filtered gapminder dataset -  with a subset of data on global development from 1952 - 2007 in increments of 5 years - to capture the period between the Second World War and the Global Financial Crisis. 

**Your task: Explore the data and visualise it in both static and animated ways, providing answers and solutions to 7 questions/tasks below.**

## Get the necessary packages
First, start with installing the relevant packages 'tidyverse', 'gganimate', and 'gapminder'. **I also used the 'scales' package which I loaded here as well**.

```{r libraries, echo = FALSE}
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("av")
# install.packages("gapminder")
library(tidyverse)
library(gganimate)
library(gifski)
library(av)
library(gapminder)
library(scales)
```

## Look at the data and tackle the tasks
First, see which specific years are actually represented in the dataset and what variables are being recorded for each country. Note that when you run the cell below, Rmarkdown will give you two results - one for each line - that you can flip between.
```{r}
str(gapminder)
unique(gapminder$year)
head(gapminder)
```

The dataset contains information on each country in the sampled year, its continent, life expectancy, population, and GDP per capita.

Let's plot all the countries in 1952.
```{r 1957}
theme_set(theme_bw())  # set theme to white background for better visibility

ggplot(subset(gapminder, year == 1952), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

We see an interesting spread with an outlier to the right. Answer the following questions, please:


1. *Why does it make sense to have a log10 scale on x axis?*

A logarithmic axis linearises exponential growth. It more clearly shows the development.

2. *Who is the outlier (the richest country in 1952 - far right on x axis)?* 

The answer is **Kuwait**. The country had a GDP per capita of 108382.3529 in 1952. I got the answer by isolating the 1952 information about country and GDP per capita into a new dataframe, `gdpPercap_1952` and then ordering this in a decreasing order by GDP pr. capita.
```{r richest country 1952}
gdpPercap_1952 <- gapminder %>% 
  filter(year == 1952) %>% 
  select(country, gdpPercap)
gdpPercap_1952[order(gdpPercap_1952$gdpPercap,decreasing=TRUE),]
```

Next, you can generate a similar plot for 2007 and compare the differences
```{r 2007}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() 
```

...

The black bubbles are a bit hard to read, the comparison would be easier with a bit more visual differentiation.

Tasks:

3. *Differentiate the **continents** by color, and fix the axis labels and units to be more legible (**Hint**: the 2.50e+08 is so called "scientific notation", which you might want to eliminate)* 

I added another argument in the aesthetic mapping for color and tied it to continent. Then I used the scale function to change the axis labels and units to no longer be scientific notations.

```{r 2007 colour}
ggplot(subset(gapminder, year == 2007), aes(gdpPercap, lifeExp, size = pop, color = continent)) +
  geom_point() +
  scale_x_log10(labels = label_number())+
  scale_size_continuous(labels = label_number())
```

4. *What are the five richest countries in the world in 2007?* 

The world's five richest countries in 2007 were **Norway, Kuwait, Singapore, United States, and Ireland**. I did the same thing as with 1952 and just counted the five first countries (there is probably a better way to do this, though).

```{r richest countries 2007}
gdpPercap_2007 <- gapminder %>% 
  filter(year == 2007) %>% 
  select(country, gdpPercap)
gdpPercap_2007[order(gdpPercap_2007$gdpPercap,decreasing=TRUE),]
```

##  Make it move!

The comparison would be easier if we had the two graphs together, animated. We have a lovely tool in R to do this: the `gganimate` package. Beware that there may be other packages your operating system needs in order to glue interim images into an animation or video. Read the messages when installing the package.

Also, there are *two* ways of animating the gapminder ggplot.

### Option 1: Animate using transition_states() 

The first step is to create the object-to-be-animated
```{r anim1 start}
anim <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10()  # convert x to log scale
anim
```

...

This plot collates all the points across time. The next step is to split it into years and animate it. This may take some time, depending on the processing power of your computer (and other things you are asking it to do). Beware that the animation might appear in the bottom right 'Viewer' pane, not in this rmd preview. You need to `knit` the document to get the visual inside an html file.

```{r anim1}
anim + transition_states(year, 
                      transition_length = 1,
                      state_length = 1)

```
...

Notice how the animation moves jerkily, 'jumping' from one year to the next 12 times in total. This is a bit clunky, which is why it's good we have another option. 


### Option 2 Animate using transition_time()
This option smoothes the transition between different 'frames', because it interpolates and adds transitional years where there are gaps in the timeseries data.

```{r anim2}
anim2 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop)) +
  geom_point() +
  scale_x_log10() + # convert x to log scale
  transition_time(year)
anim2
```

The much smoother movement in Option 2 will be much more noticeable if you add a title to the chart, that will page through the years corresponding to each frame.

Now, choose one of the animation options and get it to work. You may need to troubleshoot your installation of `gganimate` and other packages 


5. *Can you add a title to one or both of the animations above that will change* 
 *in sync with the animation?* *(**Hint**: search labeling for `transition_states()` and `transition_time()` functions respectively)* 
 
 **See below**.

6. *Can you make the axes' labels and units more readable? Consider expanding the abbreviated labels as well as the scientific notation in the legend and x axis to whole numbers.* 

I combined question 5 and 6:
```{r anim3}
anim3 <- ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = continent)) +
  geom_point() +
  labs(title = "Year: {as.integer(frame_time)}", 
       x = "GDP per capita", 
       y = "Life expectancy (years)", 
       size = "Population", 
       colour = "Continent") +
  scale_x_log10(labels = label_number())+
  scale_size_continuous(labels = label_number())+
  transition_time(year)
anim3
```


7. *Come up with a question you want to answer using the gapminder data and write it down. Then, create a data visualisation that answers the question and explain how your visualization answers the question. (Example: you wish to see what was mean life expectancy across the continents in the year you were born versus your parents' birth years). [Hint: if you wish to have more data than is in the filtered gapminder, you can load either the `gapminder_unfiltered` dataset and download more at https://www.gapminder.org/data/ ]*

I attempted to answer the following question: **_What was the mean life expectancy across continents the year I turned 1 (1997) versus the year my father turned 31 (1967)?_**

I tried out a few different ways of answering and illustrating this.
Firstly, I created a new dataframe `mean_lifeExp6797` in which I only included information about continent, year and life expectancy. I then grouped by continent and year and summarised the life expectancy column into a new column showing the mean values per year per continent. I then filtered the dataframe to only show the years relevant to answering my question, namely 1967 and 1997.
```{r mean life expectancy 1967-1997 table}
mean_lifeExp6797 <- gapminder %>% 
  select(continent,year, lifeExp) %>% 
  group_by(continent,year) %>% 
  summarise(mean_life_exp = mean(lifeExp)) %>% 
  filter(year==1967|year==1997) 
mean_lifeExp6797
```
Then I tried plotting the information in two different ways. The most tricky part of this was getting the plots to show years as separate values and not as a scale. I fixed this by making a new dataframe, `mean_lifeExp_6797_plotting` with a column where the values in the 'year' column had been converted into characters.

I then made a barplot with this new dataframe comparing the mean life expectancy of all continents in 1967 versus 1997:
```{r mean life expectancy 1967-1997 plot 1}
mean_lifeExp6797_plotting <- mean_lifeExp6797 %>% 
  mutate(year_character = as.character(year))

ggplot(mean_lifeExp6797_plotting, aes(x = year_character, y = mean_life_exp, fill = continent)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean life expectancy per continent in 1967 versus 1997",
       x = "Year",
       y = "Mean life expectancy (years)")
```

Finally, I used _faceing_ to create a collection of small barplots, one per continent, comparing mean life expectancy in 1967 versus 1997. By making both types of barplot, you can compare the development of mean life expectancy both within the resepective continents as well as between continents of the world.

```{r mean life expectancy 1967-1997 plot 2}
ggplot(mean_lifeExp6797_plotting, aes(x = year_character, y = mean_life_exp, fill = year_character)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title="Mean life expectancy per continent in 1967 versus 1997",
       x = "Continent",
       y = "Mean life expectancy (years)") +
  facet_wrap(~ continent)
```

